Effects of Reynolds Number on the Energy Conversion and Near-Wake Dynamics of a High Solidity Vertical-Axis Cross-Flow Turbine

### 2016-01-14 paper.tex
Wosnik edits part 1

% Abstract (Less than 200 words)
\abstract{Experiments were performed with a large laboratory-scale high solidity
    cross-flow turbine to investigate Reynolds number effects on performance and
    wake characteristics, to establish scale thresholds for physical and numerical
    modeling of individual devices and arrays. It was demonstrated that the
    performance of the cross-flow turbine becomes essentially $Re$-independent at a
    Reynolds number based on turbine diameter of $Re_D \approx 10^6$, or an
    approximate average Reynolds number based on blade chord of $Re_c \approx 2
    \times 10^5$. A simple model that calculates peak torque coefficient from static
    foil data and cross-flow turbine kinematics was shown to be a reasonable
    estimator for predicting Reynolds number dependence of an actual cross-flow
    turbine operating under dynamic conditions. Mean velocity and turbulence
    measurements in the near-wake showed subtle differences over the range of $Re$
    investigated. However, when transport terms for the streamwise momentum and mean
    kinetic energy were calculated, a similar $Re$ threshold was revealed. These
    results imply that physical model studies of cross-flow turbines should achieve
    $Re_D \sim 10^6$ to properly approximate both the performance and wake dynamics
    of full scale devices and arrays.}

@@

Scaled physical models are often used in science and engineering to approximate
real-world systems. The principle of dynamic similarity allows for geometrically
scaled systems to be dynamically identical if certain nondimensional physical
parameters are matched. In the case of fluid systems, the most important
nondimensional parameter is often the Reynolds number, $Re$, which quantifies
the importance of inertial forces over viscous forces on the flow physics
\cite{Acheson1990}: $Re \equiv Ul/\nu$, where $U$ and $l$ are characteristic
velocity and length scales, respectively, and $\nu$ is the fluid kinematic
viscosity. The other common dynamical scale---for systems with a free
surface---is the Froude number $Fr \equiv U/\sqrt{gl}$, where $g$ is the
gravitational acceleration. Matching both Froude and Reynolds number is
difficult for the same fluid, which is illustrated by the relation $Re = l^{3/2}
g^{1/2} Fr / \nu$, since $Re$ scales linearly with $l$, or the geometric scale.
In general the geometric scale of a physical model is not necessarily the same
as its dynamical scale. As such, hereafter we will use the word ``scale'' to
refer to this dynamical scale or Reynolds number and assume Froude number is
small enough to be negligible.

@@

Blackwell et al. \cite{Blackwell1976} investigated the effects of Reynolds
number on the performance of a 2 m diameter Darrieus vertical-axis cross-flow
wind turbine with NACA 0012 blades. By varying wind tunnel speed, the turbine
was made to operate at approximately constant blade chord Reynolds number $Re_c$
ranging from $1 \times 10^5$ to $3 \times 10^5$. In this regime the turbine
power coefficient $C_P$ was shown to be sensitive to $Re_c$, with sensitivity
diminishing at the higher Reynolds numbers, especially for turbines of lower
solidity ($Nc/R$, where $N$ is the number of blades and $R$ is the turbine's
maximum radius). More recently, Polagye and Cavagnaro \cite{Polagye2013b}
observed significant Reynolds number sensitivity for a high solidity helical
cross-flow turbine, and Bravo et al. \cite{Bravo2007} observed power coefficient
of a straight-bladed VAWT to become Reynolds number independent at $Re_c \approx
4 \times 10^5$. Bachant and Wosnik \cite{Bachant2014} observed Reynolds number
independence of power coefficient at optimal tip speed ratio for a high solidity
cross-flow turbine at $Re_D \approx 10^6$ or $Re_{c,\mathrm{avg}} \approx 2
\times 10^5$.

@@

Wake velocity was measured using a Nortek Vectrino+ acoustic Doppler velocimeter
(ADV), which has a 6 mm diameter sampling volume, and sampled at 200 Hz. The
probe was mounted on an automated positioning system, also controlled by the tow
tank's main motion controller. The ADV and DAQ systems' sampling times were
synchronized by triggering the start of data acquisition via a pulse sent from
the motion controller. Additional details of the turbine and experimental setup
are described in \cite{Bachant2015-JoT}.

@@

From each set of tows, a standard time interval was set, which allowed the
turbine performance and wake to reach a quasi-periodic state. Each run was
analyzed to compute statistics over this interval, truncating the end slightly
to achieve and integer number of blade passages. Turbine shaft angular velocity
and tow carriage speed were calculated using as second order central
differencing scheme on the respective position measurements. Power and drag
coefficients are calculated as instantaneous quantities using the carriage speed
as the free stream velocity.

Wake velocity data were filtered for spurious ``spikes'' by removing datapoints
8 standard deviations or 0.9 m/s from the mean. The experimental data and code
for processing and visualization are available from
\cite{Bachant2015-RVAT-Re-dep-data}.

@@

Uncertainty was considered from a combination of systematic and random errors.
The random error was inferred from the sample standard deviation (on a
per-revolution basis) and the systematic from the sensor calibrations or
datasheets. Combining both sources of error, along with their propagation into
quantities derived from multiple measurements, followed the procedures outlined
in Coleman and Steele \cite{ColemanSteele}, described below.

@@

\begin{table}
    \centering
\begin{tabular}{cccc}
$U_\infty$ (m/s) &                $U$ &                $V$ &                $W$ \\
\hline
           $0.4$ & $1 \times 10^{-2}$ & $7 \times 10^{-3}$ & $6 \times 10^{-3}$ \\
           $0.6$ & $1 \times 10^{-2}$ & $8 \times 10^{-3}$ & $8 \times 10^{-3}$ \\
           $0.8$ & $2 \times 10^{-2}$ & $1 \times 10^{-2}$ & $1 \times 10^{-2}$ \\
           $1.0$ & $2 \times 10^{-2}$ & $1 \times 10^{-2}$ & $1 \times 10^{-2}$ \\
           $1.2$ & $2 \times 10^{-2}$ & $1 \times 10^{-2}$ & $1 \times 10^{-2}$ \\
\end{tabular}
\caption{Average expanded uncertainty estimates (with 95\% confidence) for mean
    velocity measurements at each tow speed.}

@@

Mean power and drag coefficients at $\lambda=1.9$ are plotted versus Reynolds
number in Figure~\ref{fig:perf-Re-dep}. Note that the large error bars for $C_P$
at low $Re$ are dominated by systematic error estimates for the torque
transducer, since torque values are at the lower end of the measurement range.
However, the uncertainty due to random error or repeatability remains relatively
low. There is a drastic improvement in $C_P$ with increasing Reynolds number at
lower values of Reynolds number. The power coefficient then becomes  essentially
$Re$-independent at $Re_D = 0.8 \times 10^6$, which corresponds to an
approximate average blade chord Reynolds number $Re_{c, \mathrm{ave}} = 2.1
\times 10^5$. This threshold is consistent with the behavior of the blade
boundary layer transitioning from laminar to turbulent, thereby promoting either
the suppression or reattachment of the laminar separation bubble
\cite{Lissaman1983}.

The behavior of the mean rotor drag coefficient $C_D$ is similar, though the
changes are less dramatic. This is likely due to cross-flow turbine geometry,
where increases in blade drag at lower $Re$ somewhat offsets the reduction in
lift, causing the total streamwise force to vary less than the rotor torque. The
tendency for $C_D$ to continue increasing with $Re$ may also include the effects
of an increasing Froude number, which increases free surface deformation and
wave drag during towing without increasing flow through the turbine.

>>>>

% Abstract (Less than 200 words)
\abstract{Experiments were performed with a large laboratory-scale high solidity
    cross-flow turbine to investigate Reynolds number effects on performance and
    wake characteristics, and to establish scale thresholds for physical and
    numerical modeling of individual devices and arrays. It was demonstrated that
    the performance of the cross-flow turbine becomes essentially $Re$-independent
    at a Reynolds number based on turbine diameter of $Re_D \approx 10^6$, or an
    approximate average Reynolds number based on blade chord of $Re_c \approx 2
    \times 10^5$. A simple model that calculates peak torque coefficient from static
    foil data and cross-flow turbine kinematics was shown to be a reasonable
    estimator for predicting Reynolds number dependence of an actual cross-flow
    turbine operating under dynamic conditions. Mean velocity and turbulence
    measurements in the near-wake showed subtle differences over the range of $Re$
    investigated. However, when transport terms for the streamwise momentum and mean
    kinetic energy were calculated, a similar $Re$ threshold was revealed. These
    results imply that physical model studies of cross-flow turbines should achieve
    $Re_D \sim 10^6$ to properly approximate both the performance and wake dynamics
    of full scale devices and arrays.}

@@

Scaled physical models are often used in science and engineering to approximate
real-world systems. The principle of dynamic similarity allows for geometrically
scaled systems to be dynamically identical if certain nondimensional physical
parameters are matched. In the case of fluid systems, the most important
nondimensional parameter is often the Reynolds number, $Re$, which quantifies
the importance of inertial forces over viscous forces on the flow physics
\cite{Acheson1990}: $Re \equiv Ul/\nu$, where $U$ and $l$ are characteristic
velocity and length scales, respectively, and $\nu$ is the fluid kinematic
viscosity. The other common dynamical scale---for systems with a free
surface---is the Froude number $Fr \equiv U/\sqrt{gl}$, where $g$ is the
gravitational acceleration. Matching both Froude and Reynolds number is not
possible for a scaled model in a given fluid, which is illustrated by the
relation $Re = l^{3/2} g^{1/2} Fr / \nu$, since $Re$ scales linearly with $l$,
or the geometric scale. In general the geometric scale of a physical model is
not necessarily the same as its dynamical scale. As such, hereafter we will use
the word ``scale'' to refer to this dynamical scale or Reynolds number and
assume Froude number is small enough to be negligible.

@@

Blackwell et al. \cite{Blackwell1976} investigated the effects of Reynolds
number on the performance of a 2 m diameter Darrieus vertical-axis cross-flow
wind turbine with NACA 0012 blades. By varying wind tunnel speed, the turbine
was made to operate at approximately constant blade chord Reynolds number $Re_c$
ranging from $1 \times 10^5$ to $3 \times 10^5$. In this regime the turbine
power coefficient $C_P$ was shown to be sensitive to $Re_c$, with sensitivity
diminishing at the higher Reynolds numbers, especially for turbines of lower
solidity ($Nc/R$, where $N$ is the number of blades and $R$ is the turbine's
maximum radius). More recently, Polagye and Cavagnaro \cite{Polagye2013b}
observed significant Reynolds number sensitivity for a high solidity helical
cross-flow turbine, and Bravo et al. \cite{Bravo2007} observed the power
coefficient of a straight-bladed VAWT to become Reynolds number independent at
$Re_c \approx 4 \times 10^5$. Bachant and Wosnik \cite{Bachant2014} observed
Reynolds number independence of power coefficient at optimal tip speed ratio for
a high solidity cross-flow turbine at $Re_D \approx 10^6$ or
$Re_{c,\mathrm{avg}} \approx 2 \times 10^5$.

@@

Wake velocity was measured using a Nortek Vectrino+ acoustic Doppler velocimeter
(ADV), which has an approximately 6 mm diameter sampling volume, and sampled at
200 Hz. The probe was mounted on an automated positioning system, also
controlled by the tow tank's main motion controller. The ADV and DAQ systems'
sampling times were synchronized by triggering the start of data acquisition via
a pulse sent from the motion controller. Additional details of the turbine and
experimental setup are described in \cite{Bachant2015-JoT}.

@@

From each set of tows, a standard time interval was set, which allowed the
turbine performance and wake to reach a quasi-periodic state. Each run was
analyzed to compute statistics over this interval, truncating the end slightly
to achieve an integer number of blade passages. Turbine shaft angular velocity
and tow carriage speed were calculated using a second order central differencing
scheme on the respective position measurements. Power and drag coefficients were
calculated as instantaneous quantities using the carriage speed as the free
stream velocity.

Wake velocity data were filtered for spurious ``spikes'' by removing datapoints
8 standard deviations, or 0.9 m/s from the mean. The experimental data and code
for processing and visualization are available from
\cite{Bachant2015-RVAT-Re-dep-data}.

@@

Uncertainty was considered from a combination of systematic and random errors.
The random error was inferred from the sample standard deviation (on a
per-revolution basis) and the systematic error from the sensor calibrations or
datasheets. Combining both sources of error, along with their propagation into
quantities derived from multiple measurements, followed the procedures outlined
in Coleman and Steele \cite{ColemanSteele}, described below.

@@

\begin{table}
    \centering
\begin{tabular}{cccc}
$U_\infty$ (m/s) &          $U$ (m/s) &          $V$ (m/s) &          $W$ (m/s) \\
\hline
           $0.4$ & $1 \times 10^{-2}$ & $7 \times 10^{-3}$ & $6 \times 10^{-3}$ \\
           $0.6$ & $1 \times 10^{-2}$ & $8 \times 10^{-3}$ & $8 \times 10^{-3}$ \\
           $0.8$ & $2 \times 10^{-2}$ & $1 \times 10^{-2}$ & $1 \times 10^{-2}$ \\
           $1.0$ & $2 \times 10^{-2}$ & $1 \times 10^{-2}$ & $1 \times 10^{-2}$ \\
           $1.2$ & $2 \times 10^{-2}$ & $1 \times 10^{-2}$ & $1 \times 10^{-2}$ \\
\end{tabular}
\caption{Average expanded uncertainty estimates (with 95\% confidence) for mean
    velocity measurements at each tow speed.}

@@

Mean power and drag coefficients at $\lambda=1.9$ are plotted versus Reynolds
number in Figure~\ref{fig:perf-Re-dep}. Note that the large error bars for $C_P$
at low $Re$ are dominated by systematic error estimates for the torque
transducer, since torque values are at the lower end of its measurement range.
However, the uncertainty due to random error or repeatability remains relatively
low. There is a drastic improvement in $C_P$ with increasing Reynolds number at
lower values of Reynolds number. The power coefficient then becomes  essentially
$Re$-independent at $Re_D = 0.8 \times 10^6$, which corresponds to an
approximate average blade chord Reynolds number $Re_{c, \mathrm{ave}} = 2.1
\times 10^5$. This threshold is consistent with the behavior of the blade
boundary layer transitioning from laminar to turbulent, thereby promoting either
the suppression or reattachment of the laminar separation bubble
\cite{Lissaman1983}.

The behavior of the mean rotor drag coefficient $C_D$ is similar, though the
changes are less dramatic. This is likely due to cross-flow turbine geometry,
where increases in blade drag at lower $Re$ somewhat offsets the reduction in
lift, causing the total streamwise force to vary less than the rotor torque. The
tendency for $C_D$ to continue increasing with $Re$ may also include the effects
of an increasing Froude number (from faster tow speeds), which increases free
surface deformation and wave drag during towing without increasing flow through
the turbine.
